---
title: "Census ACS Tutorial"
author: "Jacob Armitage"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    code-fold: show
    code-overflow: wrap
    code-tools: true
    theme: cosmo
    fig-format: svg
---

```{r loading libraries}
# install.packages("pacman") optional but nice way to load multiple packages
pacman::p_load(tidyverse, tidycensus)

```

## The Census API

Obtain Census API key <https://api.census.gov/data/key_signup.html>

Once obtained and activated from your email install the API on your computer. Setting install = true will add the API key to a .Renviron file typically located in `C:/Users/yourname/Documents/.Renviron`

This only needs to be done one time

```{r Install API Key}
# install.packages("tidycensus") if not installed
# census_api_key("YOUR KEY GOES HERE", install = TRUE)
```

## Getting Census Data

Using function [`get_decennial()`](https://walker-data.com/tidycensus/reference/get_decennial.html) users can retrieve data from the 2000, 2010 or 2020 US Censuses.

Lets get the 2010 population estimates for each County in WA.

```{r Get 2010 WA Population}
#| message: false
population_WA_Counties_2010 <- get_decennial(
  geography = "county",
  year = 2010,
  state = "WA",
  variable = "P001001"
)

```

```{r}
#| echo: FALSE
head(population_WA_Counties_2010, n = 5)
 population_WA_Counties_2010 %>% summarise(mean_population = mean(value))

```
