---
title: "Census & ACS Data Tutorial"
author: "Jacob Armitage"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    code-fold: show
    code-overflow: wrap
    code-tools: false
    theme: cosmo
    fig-format: svg
editor_options: 
  chunk_output_type: console
---

Notes and examples based on [Analyzing US Census Data: Methods, Maps, and models in R](https://walker-data.com/census-r/index.html)

```{r required packages}
# install.packages("pacman") optional but nice way to install and load multiple packages at once, can also use install.packages and library separately 
pacman::p_load(tidyverse, tidycensus)

```

```{r packages for making document only}
#| echo: false
pacman::p_load(gt)
```

## The Census API

Obtain Census API key <https://api.census.gov/data/key_signup.html>

Once obtained and activated from your email install the API on your computer. Setting install = true will add the API key to a .Renviron file typically located in `C:/Users/yourname/Documents/.Renviron`

This only needs to be done one time

```{r Install API Key}
# census_api_key("YOUR KEY GOES HERE", install = TRUE)
```

## Getting Census Data

Using function [`get_decennial()`](https://walker-data.com/tidycensus/reference/get_decennial.html) users can retrieve data from the 2000, 2010 or 2020 US Censuses.

Lets get the 2010 and 2020 population estimates for each County in WA and look at 10 rows in each dataset.

```{r Get 2010 WA Population}
#| message: false
population_WA_Counties_2010 <- get_decennial(
  geography = "county",
  year = 2010,
  state = "WA",
  variable = "P001001"
)

```

```{r}
#| echo: FALSE
head(population_WA_Counties_2010, n = 10) %>% 
  gt() %>% 
  tab_options(
    table.align = "left"  # aligns the entire table left
  )

```

```{r Get 2020 WA Population}
#| message: false
population_WA_Counties_2020 <- get_decennial(
  geography = "county",
  year = 2020,
  state = "WA",
  variable = "P1_001N"
)

```

```{r}
#| echo: FALSE
head(population_WA_Counties_2020, n = 10) %>% 
  gt() %>% 
  tab_options(
    table.align = "left"  # aligns the entire table left
  )

```

## Getting ACS Data

Using function [`get_ACS()`](https://walker-data.com/tidycensus/reference/get_acs.html) users can retrieve data from single year or 5 year ACS tables. This function defaults to the most recent 5 year ACS data if not otherwise specified. Unlike the Decennial data variables from ACS tables contain both an estimated value and Margin of Error. Users can specify variable names and get_acs will auto-detect what table to look for them in. Alternatively, users can also specify a table name to get all the variables in that table.

Lets get table B01001 which gives counts of sex by age group for the most recent 5 year ACS data release in 2023.

```{r Get age & sex counts by county}
#| message: false
wa_counties_agesex <- get_acs(
  geography = "county",
  state = "WA",
  table = "B01001",
  year = 2023
)
```

```{r}
#| echo: FALSE
head(wa_counties_agesex, n = 10) %>% 
  gt() %>% 
  tab_options(
    table.align = "left"  # aligns the entire table left
  )

```

## Geography levels

The geography aggregation level available differs between get_acs() and get_decennial() with the latter going down to the Census block-level and the former having the smallest enumeration unit of block group at the 5 year level or for populations above 65,000 for the 1 year ACS data. Other political and statistical enumeration levels are available as well, see section 2.2 in the book for a table.

Lets get population counts at the 2020 Census Tract level for Thurston County, WA.

```{r Get Thurston Pop}
#| message: false
thurston_tract_pop_2020 <- get_decennial(
  geography = "tract",
  year = 2020,
  state = "WA",
  county = "Thurston",
  variable = "P1_001N"
)
```

```{r}
#| echo: FALSE
hist(thurston_tract_pop_2020$value, 
     col = "#3E938BFF", 
     main = "Thurston County Census Tract Population Counts",
     xlab = "")

```

## Searching for Variables

There are thousands of variables ids across census tables and summary files. The [load_variables()](https://walker-data.com/tidycensus/reference/load_variables.html) function obtains a dataset of variables from the Census website and formats it for searching. The function takes two arguments, `year` which is the year of the Census or end year of the ACS, and `dataset`. For 2020 data set "pl" represents the PL-94171 Redistricting dataset. 

```{r}

```


To get variables ACS users need to specify 
